---
title: "Exercise 6"
author: "Put your name here"
date: "Put the date here"
output: html_document
---

# Task 1: Reflection

Put your reflection here


# Task 2: Visualizing uncertainty with gapminder

For this exercise you'll revisit Hans Rosling's gapminder data on health and wealth. Install the **gapminder** package if you haven't already. Also install the **ggridges** and **gghalves** packages.

```{r load-libraries-data, warning=FALSE, message=FALSE}
library(tidyverse)  # For ggplot, dplyr, and friends
library(gapminder)  # For health/wealth data
library(ggridges)   # For ridge plots
library(gghalves)   # For half geoms
```

## Histograms

Make a histogram of *logged* GDP per capita for 1997 only, across all five continents. 

```{r filter-gapminder-1997}
# Create a filtered dataset
```

```{r plot-histogram}
# Add a histogram here. You'll probably want to fill and facet.
# You'll also need to add scale_x_log10()
```


## Densities

Make a ridge plot of life expectancy over time, from 1952 to 2007. You'll need to use the full gapminder data, not the 1997-only data.

```{r plot-ridgeplot}
# Add a ridge plot here
```


## Boxes, violins, and friends

Make a filtered dataset that selects data from only 2007 *and* removes Oceania. Show the distribution of *logged* GDP per capita across the *four* continents using some combination of boxplots and/or violin plots and/or strip plots, either overlaid on top of each other, or using their `geom_half_*()` counterparts from **gghalves**.

```{r filter-gapminder-2007-sans-oceania}
# Create a filtered dataset
```

```{r plot-distribution}
# Add a plot here. You'll want to use either scale_x_log10() or scale_y_log10(),
# depending on which direction your boxplots/violin plots/etc. go
```
