---
title: "Annotations"
linktitle: "9: Annotations"
date: "2020-05-21"
class_date: "2020-05-21"
output:
  blogdown::html_page:
    toc: true
menu:
  example:
    parent: Examples
    weight: 9
type: docs
weight: 9
editor_options: 
  chunk_output_type: console
---

For this example, we're again going to use cross-national data from the [World Bank's Open Data portal](https://data.worldbank.org/). We'll download the data with the [**WDI** package](https://cran.r-project.org/web/packages/WDI/index.html).

If you want to skip the data downloading, you can download the data below (you'll likely need to right click and choose "Save Link As…"):

- [<i class="fas fa-file-csv"></i> `wdi_co2.csv`](/data/wdi_co2.csv)


## Live coding example

<div class="embed-responsive embed-responsive-16by9">
<iframe class="embed-responsive-item" src="https://www.youtube.com/embed/gMSMsOy7KF0" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</div>


## Complete code

*(This is a slightly cleaned up version of the code from the video.)*

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 3.6, fig.align = "center")
set.seed(1234)
options("digits" = 2, "width" = 150)
```


### Load data

First, we load the libraries we'll be using:

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)  # For ggplot, dplyr, and friends
library(WDI)        # Get data from the World Bank
library(ggrepel)    # For non-overlapping labels

# You need to install ggtext from GitHub. Follow the instructions at 
# https://github.com/wilkelab/ggtext
library(ggtext)     # For fancier text handling
```

```{r get-wdi-fake, eval=FALSE}
indicators <- c("SP.POP.TOTL",     # Population
                "EN.ATM.CO2E.PC",  # CO2 emissions
                "NY.GDP.PCAP.KD")  # GDP per capita

wdi_co2_raw <- WDI(country = "all", indicators, extra = TRUE, 
                   start = 1995, end = 2015)
```

```{r save-wdi, eval=FALSE, include=FALSE}
write_csv(wdi_co2_raw, here::here("static", "data", "wdi_co2.csv"))
```

```{r load-data-real, include=FALSE}
wdi_co2_raw <- read_csv(here::here("static", "data", "wdi_co2.csv"))
```

Then we clean the data by removing non-country countries and renaming some of the columns.

```{r clean-data}
wdi_clean <- wdi_co2_raw %>% 
  filter(region != "Aggregates") %>% 
  select(iso2c, iso3c, country, year, 
         population = SP.POP.TOTL,
         co2_emissions = EN.ATM.CO2E.PC, 
         gdp_per_cap = NY.GDP.PCAP.KD, 
         region, income)
```

### Clean and reshape data

Next we'll do some substantial filtering and reshaping so that we can end up with the rankings of CO~2~ emissions in 1995 and 2014. I annotate as much as possible below so you can see what's happening in each step.

```{r rearrange-data}
co2_rankings <- wdi_clean %>% 
  # Get rid of smaller countries
  filter(population > 200000) %>% 
  # Only look at two years
  filter(year %in% c(1995, 2014)) %>% 
  # Get rid of all the rows that have missing values in co2_emissions
  drop_na(co2_emissions) %>% 
  # Look at each year individually and rank countries based on their emissions that year
  group_by(year) %>% 
  mutate(ranking = rank(co2_emissions)) %>% 
  ungroup() %>% 
  # Only select a handful of columns, mostly just the newly created "ranking"
  # column and some country identifiers
  select(iso3c, country, year, region, income, ranking) %>% 
  # Right now the data is tidy and long, but we want to widen it and create
  # separate columns for emissions in 1995 and in 2014. pivot_wider() will make
  # new columns based on the existing "year" column (that's what `names_from`
  # does), and it will add "rank_" as the prefix, so that the new columns will
  # be "rank_1995" and "rank_2014". The values that go in those new columns will
  # come from the existing "ranking" column
  pivot_wider(names_from = year, names_prefix = "rank_", values_from = ranking) %>% 
  # Find the difference in ranking between 2014 and 1995
  mutate(rank_diff = rank_2014 - rank_1995) %>% 
  # Remove all rows where there's a missing value in the rank_diff column
  drop_na(rank_diff) %>% 
  # Make an indicator variable that is true of the absolute value of the
  # difference in rankings is greater than 25. 25 is arbitrary here—that just
  # felt like a big change in rankings
  mutate(big_change = ifelse(abs(rank_diff) >= 25, TRUE, FALSE)) %>% 
  # Make another indicator variable that indicates if the rank improved by a
  # lot, worsened by a lot, or didn't change much. We use the case_when()
  # function, which is like a fancy version of ifelse() that takes multiple
  # conditions. This is how it generally works:
  #
  # case_when(
  #  some_test ~ value_if_true,
  #  some_other_test ~ value_if_true,
  #  TRUE ~ value_otherwise
  #)
  mutate(better_big_change = case_when(
    rank_diff <= -25 ~ "Rank improved",
    rank_diff >= 25 ~ "Rank worsened",
    TRUE ~ "Rank changed a little"
  ))
```

Here's what that reshaped data looked like before:

```{r show-head-original}
head(wdi_clean)
```

And here's what it looks like now:

```{r show-head-new}
head(co2_rankings)
```

### Plot the data and annotate

I use IBM Plex Sans in this plot. You can [download it from Google Fonts](https://fonts.google.com/specimen/IBM+Plex+Sans).

```{r build-pretty-plot, fig.width=10, fig.height=6}
# These three functions make it so all geoms that use text, label, and
# label_repel will use IBM Plex Sans as the font. Those layers are *not*
# influenced by whatever you include in the base_family argument in something
# like theme_bw(), so ordinarily you'd need to specify the font in each
# individual annotate(geom = "text") layer or geom_label() layer, and that's
# tedious! This removes that tediousness.
update_geom_defaults("text", list(family = "IBM Plex Sans"))
update_geom_defaults("label", list(family = "IBM Plex Sans"))
update_geom_defaults("label_repel", list(family = "IBM Plex Sans"))

ggplot(co2_rankings,
       aes(x = rank_1995, y = rank_2014)) +
  # Add a reference line that goes from the bottom corner to the top corner
  annotate(geom = "segment", x = 0, xend = 175, y = 0, yend = 175) +
  # Add points and color them by the type of change in rankings
  geom_point(aes(color = better_big_change)) +
  # Add repelled labels. Only use data where big_change is TRUE. Fill them by
  # the type of change (so they match the color in geom_point() above) and use
  # white text
  geom_label_repel(data = filter(co2_rankings, big_change == TRUE),
                   aes(label = country, fill = better_big_change),
                   color = "white") +
  # Add notes about what the outliers mean in the bottom left and top right
  # corners. These are italicized and light grey. The text in the bottom corner
  # is justified to the right with hjust = 1, and the text in the top corner is
  # justified to the left with hjust = 0
  annotate(geom = "text", x = 170, y = 6, label = "Outliers improving", 
           fontface = "italic", hjust = 1, color = "grey50") +
  annotate(geom = "text", x = 2, y = 170, label = "Outliers worsening", 
           fontface = "italic", hjust = 0, color = "grey50") +
  # Add mostly transparent rectangles in the bottom right and top left corners
  annotate(geom = "rect", xmin = 0, xmax = 25, ymin = 0, ymax = 25, 
           fill = "#2ECC40", alpha = 0.25) +
  annotate(geom = "rect", xmin = 150, xmax = 175, ymin = 150, ymax = 175, 
           fill = "#FF851B", alpha = 0.25) +
  # Add text to define what the rectangles abovee actually mean. The \n in
  # "highest\nemitters" will put a line break in the label
  annotate(geom = "text", x = 40, y = 6, label = "Lowest emitters", 
           hjust = 0, color = "#2ECC40") +
  annotate(geom = "text", x = 162.5, y = 135, label = "Highest\nemitters", 
           hjust = 0.5, vjust = 1, lineheight = 1, color = "#FF851B") +
  # Add arrows between the text and the rectangles. These use the segment geom,
  # and the arrows are added with the arrow() function, which lets us define the
  # angle of the arrowhead and the length of the arrowhead pieces. Here we use
  # 0.5 lines, which is a unit of measurement that ggplot uses internally (think
  # of how many lines of text fit in the plot). We could also use unit(1, "cm")
  # or unit(0.25, "in") or anything else
  annotate(geom = "segment", x = 38, xend = 20, y = 6, yend = 6, color = "#2ECC40", 
           arrow = arrow(angle = 15, length = unit(0.5, "lines"))) +
  annotate(geom = "segment", x = 162.5, xend = 162.5, y = 140, yend = 155, color = "#FF851B", 
           arrow = arrow(angle = 15, length = unit(0.5, "lines"))) +
  # Use three different colors for the points
  scale_color_manual(values = c("grey50", "#0074D9", "#FF4136")) +
  # Use two different colors for the filled labels. There are no grey labels, so
  # we don't have to specify that color
  scale_fill_manual(values = c("#0074D9", "#FF4136")) +
  # Make the x and y axes expand all the way to the edges of the plot area and
  # add breaks every 25 units from 0 to 175
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 175, 25)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 175, 25)) +
  # Add labels! There are a couple fancy things here.
  # 1. In the title we wrap the 2 of CO2 in the HTML <sub></sub> tag so that the
  #    number gets subscripted. The only way this will actually get parsed as 
  #    HTML is if we tell the plot.title to use element_markdown() in the 
  #    theme() function, and element_markdown() comes from the ggtext package.
  # 2. In the subtitle we bold the two words **improved** and **worsened** using
  #    Markdown asterisks. We also wrap these words with HTML span tags with 
  #    inline CSS to specify the color of the text. Like the title, this will 
  #    only be processed and parsed as HTML and Markdown if we tell the p
  #    lot.subtitle to use element_markdown() in the theme() function.
  labs(x = "Rank in 1995", y = "Rank in 2014",
       title = "Changes in CO<sub>2</sub> emission rankings between 1995 and 2014",
       subtitle = "Countries that <span style='color: #0074D9'>**improved**</span> or <span style='color: #FF4136'>**worsened**</span> more than 25 positions in the rankings highlighted",
       caption = "Source: The World Bank.\nCountries with populations of less than 200,000 excluded.") +
  # Turn off the legends for color and fill, since the subtitle includes that
  guides(color = FALSE, fill = FALSE) +
  # Use theme_bw() with IBM Plex Sans
  theme_bw(base_family = "IBM Plex Sans") +
  # Tell the title and subtitle to be treated as Markdown/HTML, make the title
  # 1.6x the size of the base font, and make the subtitle 1.3x the size of the
  # base font. Also add a little larger margin on the right of the plot so that
  # the 175 doesn't get cut off.
  theme(plot.title = element_markdown(face = "bold", size = rel(1.6)),
        plot.subtitle = element_markdown(size = rel(1.3)),
        plot.margin = unit(c(0.5, 1, 0.5, 0.5), units = "lines"))
```
