---
title: "Text"
linktitle: "13: Text"
date: "2020-05-28"
class_date: "2020-05-28"
output:
  blogdown::html_page:
    toc: true
    toc_depth: 4
menu:
  example:
    parent: Examples
    weight: 13
type: docs
weight: 13
editor_options: 
  chunk_output_type: console
---

For this example, we're going to use the text of *Little Women* by Louisa May Alcott and four Shakespearean tragedies (*Romeo and Juliet*, *King Lear*, *Macbeth*, and *Hamlet*) to explore how to do some basic text visualization.

You can follow along if you want, but **don't feel like you have too**. This is mostly just to give you a taste of different methods for visualizing text. It's by no means comprehensive, but it is well annotated and commented and should (hopefully) be easy to follow.

If you want to play with part-of-speech tagging, you can download an already-tagged version of *Little Women* here (you'll likely need to right click and choose "Save Link As…"):

- [<i class="fas fa-file-csv"></i> `little_women_tagged.csv`](/data/little_women_tagged.csv)

If you want to see other examples of text visualizations with the **tidytext** package, check out some of these:

- <i class="fas fa-external-link-square-alt"></i> [Harry Potter Sentiment Analysis for Beginners](https://rstudio-pubs-static.s3.amazonaws.com/300624_8260952d1f0346969e65f41a97006bf5.html) (this uses [the **harrypotter** package](https://github.com/bradleyboehmke/harrypotter), which you can install from GitHub (not from CRAN))
- <i class="fas fa-external-link-square-alt"></i> Peer Christensen ["Fair is foul, and foul is fair: a tidytext sentiment analysis of Shakespeare’s tragedies"](https://peerchristensen.netlify.app/post/fair-is-foul-and-foul-is-fair-a-tidytext-entiment-analysis-of-shakespeare-s-tragedies/)
- <i class="fas fa-external-link-square-alt"></i> ["Tidy text, parts of speech, and unique words in the Bible"](https://www.andrewheiss.com/blog/2018/12/26/tidytext-pos-john/)
- <i class="fas fa-external-link-square-alt"></i> ["Tidy text, parts of speech, and unique words in the Qur'an"](https://www.andrewheiss.com/blog/2018/12/28/tidytext-pos-arabic/)

## Live coding example

<div class="embed-responsive embed-responsive-16by9">
<iframe class="embed-responsive-item" src="https://www.youtube.com/embed/YeyZp8Dw55g" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</div>


## Complete code

*(This is a highly cleaned up version of the code from the video.)*

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.2, fig.align = "center")
set.seed(1234)
options("digits" = 2, "width" = 150)
```

### Get data

First, as always, we'll load the libraries we'll be using:

```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)   # For ggplot, dplyr, etc.
library(tidytext)    # For neat text things
library(gutenbergr)  # For downloading books from Project Gutenberg
```

We're going to use the **gutenbergr** package to download some books directly from Project Gutenberg. The IDs for these books come from the URLs at their website. For instance, [*Little Women* is book #514](https://www.gutenberg.org/ebooks/514). We'll store these books as `*_raw* and then clean them up later.

```{r get-text-fake, eval=FALSE}
# 514 Little Women
little_women_raw <- gutenberg_download(514, meta_fields = "title")

# 1524 - Hamlet
# 1532 - King Lear
# 1533 - Macbeth
# 1513 - Romeo and Juliet
tragedies_raw <- gutenberg_download(c(1524, 1532, 1533, 1513),
                                    meta_fields = "title")
```

```{r save_text, include=FALSE, eval=FALSE}
write_csv(little_women_raw, here::here("static", "data", "little_women_raw.csv"))
write_csv(tragedies_raw, here::here("static", "data", "tragedies_raw.csv"))
```

```{r load-saved-text, include=FALSE}
little_women_raw <- read_csv(here::here("static", "data", "little_women_raw.csv"))
tragedies_raw <- read_csv(here::here("static", "data", "tragedies_raw.csv"))
```

If you won't want to redownload the books every time you knit (you don't), you can do the same trick we've used for [WDI](/example/08-example/) and [FRED data](/example/11-example/). Put the actual code for getting the books in a chunk with `eval=FALSE` on it and run it manually in RStudio when you want to get the data. Then you can write the downloaded data as a CSV file, and then load it invisibly from the CSV file when you knit:

````text
I first download data from Project Gutenberg:

```{r get-book, eval=FALSE}`r ''`
books_raw <- gutenberg_download(...)

write_csv(books_raw, "data/books_raw.csv")
```

```{r load-book-data-real, include=FALSE}`r ''`
books_raw <- read_csv("data/books_raw.csv")
```
````

### Clean data

The data you get from Project Gutenberg comes in a tidy format, with a column for the book id, a column for the title, and a column for text. Sometimes this text column will be divided by lines in the book; sometimes it might be an entire page or paragraph or chapter. It all depends on how the book is formatted at Project Gutenberg.

Here's what the start of our `little_women_raw` data looks like:

```{r show-lw-start}
head(little_women_raw)
```

If we look at the data in RStudio, we can see that the actual book doesn't start until row 70 (the first 69 rows are the table of contents and other parts of the front matter).

It would be nice if we had a column that indicated what chapter each line is in, since we could then group by chapter and look at patterns within chapters. Since the data doesn't come with a chapter column, we have to make one ourselves using a fun little trick. Each chapter in the book starts with "CHAPTER ONE" or "CHAPTER TWO", with "chapter" in ALL CAPS. We can make a variable named `chapter_start` that will be true if a line starts with "CHAPTER" and false if not. Then we can use the `cumsum()` function to take the cumulative sum of this column, which will increment up one number ever time there's a new chapter, thus creating a helpful chapter column.

```{r clean-data-lw}
# Clean up Little Women
little_women <- little_women_raw %>% 
  # The actual book doesn't start until line 70
  slice(70:n()) %>% 
  # Get rid of rows where text is missing
  drop_na(text) %>% 
  # Chapters start with CHAPTER X, so mark if each row is a chapter start
  # cumsum() calculates the cumulative sum, so it'll increase every time there's
  # a new chapter and automatically make chapter numbers
  mutate(chapter_start = str_detect(text, "^CHAPTER"),
         chapter_number = cumsum(chapter_start)) %>% 
  # Get rid of these columns
  select(-gutenberg_id, -title, -chapter_start)

head(little_women)
```

The data from Shakespeare is similarly messy, with just three columns:

```{r show-shakespeare-raw}
head(tragedies_raw)
```

The initial text sometimes isn't the actual text of the book. If you look at the beginning of *Hamlet*, for instance, there's a bunch of introductory stuff from editors and transcribers. In real life, we'd want to find a systematic way to get rid of that (perhaps by looking at how many introductory rows there are in each of the four plays and removing those rows), but for now, we'll just live with it and pretend Shakespeare wrote these notes. `r emo::ji("shrug")`

We could also figure out a systematic way to indicate acts and scenes, but that's tricky, so we won't for this example. ([This guy did though!](https://peerchristensen.netlify.app/post/fair-is-foul-and-foul-is-fair-a-tidytext-entiment-analysis-of-shakespeare-s-tragedies/))

Now that we have tidy text data, let's do stuff with it!


### Tokens and word counts

#### Single words

One way we can visualize text is to look at word frequencies and find the most common words. This is even more important when looking across documents. 

Right now the text we have is tidy, but it is based on lines of text, not words. In order to count words correctly, we need each token (or text element, whether it be a word or bigram or paragraph or whatever) to be in its own row. The `unnest_tokens()` functions from **tidytext** does this for us. The first argument is the name of the column we want to create; the second argument is the name of the column we want to split into tokens.

Let's just work with the Shakespeare tragedies:

```{r tidy-shakespeare-1}
tragedies_words <- tragedies_raw %>% 
  drop_na(text) %>% 
  unnest_tokens(word, text)

head(tragedies_words)
```

Now that we have words, we can filter and count the words. Here's what's happening in this next chunk:

- We use `anti_join()` to remove all common stop words like "a" and "the" that are listed in the `stop_words` dataset that is loaded when you load **tidytext**
- We count how many times each word appears in each title/play
- We only keep the top 15 words

```{r top-15-shakespeare, message=FALSE, warning=FALSE}
top_words_tragedies <- tragedies_words %>% 
  # Remove stop words
  anti_join(stop_words) %>% 
  # Get rid of old timey words and stage directions
  filter(!(word %in% c("thou", "thy", "haue", "thee", 
                      "thine", "enter", "exeunt", "exit"))) %>% 
  # Count all the words in each play
  count(title, word, sort = TRUE) %>% 
  # Keep top 15 in each play
  group_by(title) %>% 
  top_n(15) %>% 
  ungroup() %>% 
  # Make the words an ordered factor so they plot in order
  mutate(word = fct_inorder(word))
top_words_tragedies
```

Now we can plot these results, facetting and filling by title:

```{r plot-top-words-tragedies}
ggplot(top_words_tragedies, aes(y = fct_rev(word), x = n, fill = title)) + 
  geom_col() + 
  guides(fill = FALSE) +
  labs(y = "Count", x = NULL, 
       title = "15 most frequent words in four Shakespearean tragedies") +
  facet_wrap(vars(title), scales = "free_y") +
  theme_bw()
```

These results aren't terribly surprising. "lear" is the most common word in *King Lear*, "macbeth" is the most common word in *Macbeth*, and so on. But the results are still really neat! This is a wordcloud for grownups!

(Sharp-eyed readers will notice that the words aren't actually in perfect order! That's because some common words are repeated across the plays, like "lord" and "sir". However, each category in a factor can only have one possible position in the orer, so because "lord" is the second most common word in *Hamlet* it also appears as #2 in *Macbeth* and *King Lear*. You can fix this with the `reorder_within()` function in **tidytext**—see [Julia Silge's tutorial here](https://juliasilge.com/blog/reorder-within/) for how to use it.)

#### Bigrams

We can also look at pairs of words instead of single words. To do this, we need to change a couple arguments in `unnest_tokens()`, but otherwise everything else stays the same. In order to remove stopwords, we need to split the bigram column into two columns (`word1` and `word2`) with `separate()`, filter each of those columns, and then combine the word columns back together as `bigram` with `unite()`

```{r top-bigrams}
tragedies_bigrams <- tragedies_raw %>% 
  drop_na(text) %>% 
  # n = 2 here means bigrams. We could also make trigrams (n = 3) or any type of n-gram
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  # Split the bigrams into two words so we can remove stopwords
  separate(bigram, c("word1", "word2"), sep = " ") %>% 
  filter(!word1 %in% stop_words$word,
         !word2 %in% stop_words$word) %>% 
  filter(!word1 %in% c("thou", "thy", "thine", "enter", "exeunt", "exit"),
         !word2 %in% c("thou", "thy", "thine", "enter", "exeunt", "exit")) %>% 
  # Put the two word columns back together
  unite(bigram, word1, word2, sep = " ")
tragedies_bigrams

top_bigrams <- tragedies_bigrams %>% 
  # Count all the bigrams in each play
  count(title, bigram, sort = TRUE) %>% 
  # Keep top 15 in each play
  group_by(title) %>% 
  top_n(15) %>% 
  ungroup() %>% 
  # Make the bigrams an ordered factor so they plot in order
  mutate(bigram = fct_inorder(bigram))

ggplot(top_bigrams, aes(y = fct_rev(bigram), x = n, fill = title)) + 
  geom_col() + 
  guides(fill = FALSE) +
  labs(y = "Count", x = NULL, 
       title = "15 most frequent bigrams in four Shakespearean tragedies") +
  facet_wrap(vars(title), scales = "free") +
  theme_bw()
```

There are some neat trends here. "Lord Hamlet" is the most common pair of words in *Hamlet* (not surprisingly), but in Macbeth the repeated "knock knock" (the first non-name repeated pair) is a well-known plot point and reoccurring symbolic theme throughout the play.

### Bigrams and probability

We can replicate the ["She Giggles, He Gallops"](https://pudding.cool/2017/08/screen-direction/) idea by counting the bigrams that match "he X" and "she X". 

The log ratio idea shows how much more likely a word is compared to its counterpart (so "he that" is about 5 more likely to appear than "she that". In this graph, I replaced the x-axis labels with "2x" and "4x", but without those, you get numbers like 1, 2, and 3 (or -1, -2, -3)). To convert those logged ratio numbers into the multiplicative version (i.e. 2x instead of 1), raise 2 to the power of the log ratio. If the log ratio is 3, the human-readable version is $2^3$, or 8 times.

```{r example-logs}
# Take the log of 8:
log2(8)

# Reverse log of 3:
2^3
```

The only text wizardry here is tokenizing the words. Pretty much the rest of all this code is just **dplyr** mutating, filtering, and counting:

```{r bigrams-he-she}
pronouns <- c("he", "she")

bigram_he_she_counts <- tragedies_raw %>%
  drop_na(text) %>% 
  # Split into bigrams
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  # Find counts of bigrams
  count(bigram, sort = TRUE) %>%
  # Split the bigram column into two columns
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  # Only choose rows where the first word is he or she
  filter(word1 %in% pronouns) %>%
  count(word1, word2, wt = n, sort = TRUE) %>% 
  rename(total = n)

word_ratios <- bigram_he_she_counts %>%
  # Look at each of the second words
  group_by(word2) %>%
  # Only choose rows where the second word appears more than 10 times
  filter(sum(total) > 10) %>%
  ungroup() %>%
  # Spread out the word1 column so that there's a column named "he" and one named "she"
  spread(word1, total, fill = 0) %>%
  # Add 1 to each number so that logs work (just in case any are zero)
  mutate_if(is.numeric, ~(. + 1) / sum(. + 1)) %>%
  # Create a new column that is the logged ratio of the she counts to he counts
  mutate(logratio = log2(she / he)) %>%
  # Sort by that ratio
  arrange(desc(logratio))

# Rearrange this data so it's plottable
plot_word_ratios <- word_ratios %>%
  # This gets the words in the right order---we take the absolute value, select
  # only rows where the log ratio is bigger than 0, and then take the top 15 words
  mutate(abslogratio = abs(logratio)) %>%
  group_by(logratio < 0) %>%
  top_n(15, abslogratio) %>%
  ungroup() %>%
  mutate(word = reorder(word2, logratio)) 

# Finally we plot this
ggplot(plot_word_ratios, aes(y = word, x = logratio, color = logratio < 0)) +
  geom_segment(aes(y = word, yend = word,
                   x = 0, xend = logratio), 
               size = 1.1, alpha = 0.6) +
  geom_point(size = 3.5) +
  labs(x = "How much more/less likely", y = NULL) +
  scale_color_discrete(name = "", labels = c("More 'she'", "More 'he'")) +
  scale_x_continuous(breaks = seq(-3, 3),
                     labels = c("8x", "4x", "2x",
                                "Same", "2x", "4x", "8x")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Shakespeare doesn't use a lot of fancy verbs in his plays, so we're left with incredibly common verbs like "should" and "comes" and "was". Oh well.


### Term frequency-inverse document frequency (tf-idf)

We can determine which words are the most unique for each book/document in our corpus using by calculating the tf-idf (term frequency-inverse document frequency) score for each term. The tf-idf is the product of the term frequency and the inverse document frequency:

$$
\begin{aligned}
tf(\text{term}) &= \frac{n_{\text{term}}}{n_{\text{terms in document}}} \\
idf(\text{term}) &= \ln{\left(\frac{n_{\text{documents}}}{n_{\text{documents containing term}}}\right)} \\
tf\text{-}idf(\text{term}) &= tf(\text{term}) \times idf(\text{term})
\end{aligned}
$$

Fortunately you don't need to remember that formula. The `bind_tf_idf()` function will calculate this for you. Remember, the higher the tf-idf number, the more unique the term is in the document, but these numbers are meaningless and unitless—you can't convert them to a percentage or anything.

Here are the most unique words in these four tragedies, compared to all the tragedies:

```{r tf-idf-tragedies, message=FALSE}
tragedy_words <- tragedies_raw %>% 
  drop_na() %>% 
  # Split into word tokens
  unnest_tokens(word, text) %>% 
  # Remove stop words and old timey words
  anti_join(stop_words) %>% 
  filter(!word %in% c("thou", "thy", "haue", "thee", 
                      "thine", "enter", "exeunt", "exit")) %>% 
  count(title, word, sort = TRUE)

# Add the tf-idf values to the counts
tragedy_tf_idf <- tragedy_words %>% 
  bind_tf_idf(word, title, n)

# Get the top 10 uniquest words
tragedy_tf_idf_plot <- tragedy_tf_idf %>% 
  arrange(desc(tf_idf)) %>% 
  group_by(title) %>% 
  top_n(10) %>% 
  ungroup() %>% 
  mutate(word = fct_inorder(word))

ggplot(tragedy_tf_idf_plot, 
       aes(y = fct_rev(word), x = tf_idf, fill = title)) +
  geom_col() +
  guides(fill = FALSE) +
  labs(x = "tf-idf", y = NULL) +
  facet_wrap(~ title, scales = "free") +
  theme_bw()
```

Not surprisingly, the most unique words for each play happen to be the names of the characters in those plays.


### Sentiment analysis

In the video, I plotted the sentiment of *Little Women* across the book, but it wasn't a very interesting plot. We'll try with Shakespeare here instead. 

At its core, sentiment analysis involves looking at a big list of words for how negative or positive they are. Some sentiment dictionaries mark if a word is "negative" or "positive"; some give words a score from -3 to 3; some give different emotions like "sadness" or "anger". You can see what the different dictionaries look like with `get_sentiments()`

```{r show-dictionary}
get_sentiments("afinn")  # Scoring system
# get_sentiments("bing")  # Negative/positive
# get_sentiments("nrc")  # Specific emotions
# get_sentiments("loughran")  # Designed for financial statements; positive/negative
```

Here we split the Shakespearean tragedies into words, join a sentiment dictionary to it, and use **dplyr** data wrangling to calculate the net number positive words in each chapter. Had we used the AFINN library, we could calculate the average sentiment per chapter, since AFINN uses a scoring system instead of negative/positive labels. Or we could've used the NRC library, which has specific emotions like trust and fear.

```{r shakespeare-sentiment, message=FALSE}
tragedy_words <- tragedies_raw %>% 
  drop_na() %>% 
  # Split into word tokens
  unnest_tokens(word, text) %>% 
  # Remove stop words and old timey words
  anti_join(stop_words) %>% 
  filter(!word %in% c("thou", "thy", "haue", "thee", 
                      "thine", "enter", "exeunt", "exit"))

# Join the sentiment dictionary 
tragedy_sentiment <- tragedy_words %>% 
  inner_join(get_sentiments("bing"))
tragedy_sentiment
```

We can look at these sentiments a few different ways. First we can get a count of total positive and negative words in the four books. We can see that in all four, there are more negative words than positive ones (they're tragdies, after all):

```{r shakespeare-sentiment-plot-bars, message=FALSE, warning=FALSE}
tragedy_sentiment_plot <- tragedy_sentiment %>% 
  count(title, sentiment)

ggplot(tragedy_sentiment_plot, aes(x = sentiment, y = n, fill = title, alpha = sentiment)) +
  geom_col(position = position_dodge()) +
  scale_alpha_manual(values = c(0.5, 1)) +
  facet_wrap(vars(title)) +
  theme_bw()
```

Perhaps more usefully, we can divide each of the plays into groups of 100 lines, and then get the net sentiment of each group (number of positive words − number of negative words). By splitting the data into groups of lines, we can show a more granular view of the progression of the plot. To do this we make a column that indicates the row number, and then we use the special `%/%` operator to perform integer division, which essentially lops off the decimal point when dividing numbers: 150/100 normally is 1.5, but in integer divison, it is 1. This is a helpful trick for putting rows 1-99 in one group, then rows 100-199 in another group, etc.

```{r shakespeare-sentiment-plot-lines, message=FALSE}
tragedies_split_into_lines <- tragedy_sentiment %>% 
  # Divide lines into groups of 100
  mutate(line = row_number(),
         line_chunk = line %/% 100) %>% 
  # Get a count of postiive and negative words in each 100-line chunk in each play
  count(title, line_chunk, sentiment) %>% 
  # Convert the sentiment column into two columns named "positive" and "negative"
  pivot_wider(names_from = sentiment, values_from = n) %>% 
  # Calculate net sentiment
  mutate(sentiment = positive - negative)

ggplot(tragedies_split_into_lines,
       aes(x = line_chunk, y = sentiment, fill = sentiment)) +
  geom_col() +
  scale_fill_viridis_c(option = "magma", end = 0.9) +
  facet_wrap(vars(title), scales = "free_x") +
  theme_bw()
```

Neat. They're all really sad and negative, except for the beginning of Romeo and Juliet where the two lovers meet and fall in love. Then everyone dies later.


### Neat extra stuff

None of this stuff was in the video, but it's useful to know and see how to do it. It all generally comes from the [*Tidy Text Mining* book](https://www.tidytextmining.com/) by Julia Silge and David Robinson

#### Part of speech tagging

R has no way of knowing if words are nouns, verbs, or adjectives. You can algorithmically predict what part of speech each word is using a part-of-speech tagger, like [**spaCy**](https://spacy.io/) or [Stanford's Natural Langauge Processing (NLP) library](https://nlp.stanford.edu/). 

These are external programs that are not written in R and don't naturally communicate with R (spaCy is written in Python; Stanford's CoreNLP is written in Java). There is a helpful R package named **cleanNLP** that helps you interact with these programs from within R, whis is super helpful. **cleanNLP** also comes with its own R-only tagger so you don't need to install anything with Python or Java (however, it's not as powerful as either spaCy, which is faster, and doesn't deal with foreign languages like Arabic and Chinese like Stanford's NLP library).

You can see other examples of part-of-speech tagging (along with instructions for how to install spaCy and coreNLP) here:

- <i class="fas fa-external-link-square-alt"></i> ["Tidy text, parts of speech, and unique words in the Bible"](https://www.andrewheiss.com/blog/2018/12/26/tidytext-pos-john/)
- <i class="fas fa-external-link-square-alt"></i> ["Tidy text, parts of speech, and unique words in the Qur'an"](https://www.andrewheiss.com/blog/2018/12/28/tidytext-pos-arabic/)

Here's the general process for tagging (or "annotating") text with the **cleanNLP** package:

1. Make a dataset where one column is the id (line number, chapter number, book+chapter, etc.), and another column is the text itself.

2. Initialize the NLP tagger. You can use any of these:

    - `cnlp_init_udpipe()`: Use an R-only tagger that should work without installing anything extra (a little slower than the others, but requires no extra steps!)
    - `cnlp_init_spacy()`: Use spaCy (if you've installed it on your computer with Python)
    - `cnlp_init_corenlp()`: Use Stanford's NLP library (if you've installed it on your computer with Java) 

3. Feed the data frame from step 1 into the `cnlp_annotate()` function and wait.

4. Save the tagged data on your computer so you don't have to re-tag it every time.

Here's an example using the *Little Women* data:

```{r lw-reshape}
# For the tagger to work, each row needs to be unique, which means we need to
# combine all the text into individual chapter-based rows. This takes a little
# bit of text-wrangling with dplyr:
little_women_to_tag <- little_women %>% 
  # Group by chapter number
  group_by(chapter_number) %>% 
  # Take all the rows in each chapter and collapse them into a single cell
  nest(data = c(text)) %>% 
  ungroup() %>% 
  # Look at each individual cell full of text lines and paste them together into
  # one really long string of text per chapter
  mutate(text = map_chr(data, ~paste(.$text, collapse = " "))) %>% 
  # Get rid of this column
  select(-data)
little_women_to_tag
```

Notice how there's now a row for each chapter, and the whole chapter is contained in the `text` column. With the data in this format, we can annotate it. It takes about 3 minutes to run this on my 2016 MacBook Pro with the R-only udpipe tagger (and only 30 seconds if I use the spaCy tagger). Notice how I immediately save the tagged tokens as a CSV file after so I don't have to do it again. 

```{r nlp-example-tag, eval=FALSE}
library(cleanNLP)

# Use the built-in R-based tagger
cnlp_init_udpipe()

little_women_tagged <- cnlp_annotate(little_women_to_tag, 
                                     text_name = "text", 
                                     doc_name = "chapter_number")

write_csv(little_women_tagged$token, "little_women_tagged.csv")
```

```{r load-tagged-text, include=FALSE}
little_women_tagged <- read_csv(here::here("static", "data", "little_women_tagged.csv"))
```

```{r nlp-example-tag-spacy, eval=FALSE, include=FALSE}
write_csv(little_women_tagged$token, 
          here::here("static", "data", "little_women_tagged.csv"))

reticulate::use_python("/usr/local/bin/python3")
cnlp_init_spacy()

tictoc::tic()
little_women_tagged_spacy <- cnlp_annotate(little_women_to_tag, text_name = "text", doc_name = "chapter_number")
tictoc::toc()

little_women_tagged_spacy$token
```

Here's what the tagged text looks like:

```{r show-tagged-lw}
little_women_tagged
```

There are a bunch of new columns like `lemma` (or the base stemmed word), and `upos` and `pos` for the different parts of speech. These use the [Penn Treebank codes](https://www.ling.upenn.edu/courses/Fall_2003/ling001/penn_treebank_pos.html).

Now that everything is tagged, we can do any grouping and summarizing and filtering we want. We could find the most common verbs, or the most common nouns or proper names, for instance. Here's a fun plot that shows the proportion of mentions of the four main characters (Meg, Jo, Beth, and Amy) in each chapter.

```{r lw-props}
# Find all proper nouns
proper_nouns <- little_women_tagged %>% 
  filter(upos == "PROPN")

main_characters_by_chapter <- proper_nouns %>% 
  # Find only Meg, Jo, Beth, and Amy
  filter(lemma %in% c("Meg", "Jo", "Beth", "Amy")) %>% 
  # Group by chapter and character name
  group_by(doc_id, lemma) %>% 
  # Get the count of mentions
  summarize(n = n()) %>% 
  # Make a new column named "name" that is an ordered factor of the girls' names
  mutate(name = factor(lemma, levels = c("Meg", "Jo", "Beth", "Amy"), ordered = TRUE)) %>% 
  # Rename this so it's called chapter
  rename(chapter = doc_id) %>% 
  # Group by chapter
  group_by(chapter) %>% 
  # Calculate the proportion of each girl's mentions in each chapter
  mutate(prop = n / sum(n)) %>% 
  ungroup() %>% 
  # Make a cleaner chapter name column
  mutate(chapter_name = paste("Chapter", chapter)) %>% 
  mutate(chapter_name = fct_inorder(chapter_name))
main_characters_by_chapter
```

And here's the polished plot:

```{r lw-props-plot, fig.width=8, fig.height=5}
ggplot(main_characters_by_chapter, aes(x = prop, y = 1, fill = fct_rev(name))) + 
  geom_col(position = position_stack()) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_viridis_d(option = "plasma", end = 0.9, name = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(x = NULL, y = NULL,
       title = "Proportion of mentions of each\nLittle Woman per chapter",
       subtitle = "Jo basically dominates the last third of the book") +
  facet_wrap(vars(chapter_name), nrow = 6) +
  theme_bw(base_family = "Roboto Condensed") +
  theme(legend.position = "top",
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.background = element_rect(fill = "white"),
        legend.text = element_text(face = "bold", size = rel(1)),
        plot.title = element_text(face = "bold", hjust = 0.5, size = rel(1.7)),
        plot.subtitle = element_text(hjust = 0.5, size = rel(1.1)))
```


#### Topic modeling and fingerprinting

If you want to see some examples of topic modeling with Latent Dirichlet Allocation (LDA) or text fingerprinting based on sentence length and counts of hapax legomena ([based on this article](https://kops.uni-konstanz.de/bitstream/handle/123456789/5492/Literature_Fingerprinting.pdf)), see these examples from a previous version of this class: [topic modeling](https://datavizf18.classes.andrewheiss.com/class/11-class/#topic-modeling) and [fingerprinting](https://datavizf18.classes.andrewheiss.com/class/11-class/#fingerprinting)


#### Text features

Finally, you can use [the **textfeatures** package](https://github.com/mkearney/textfeatures) to find all sorts of interesting numeric statistics about text, like the number of exclamation points, commas, digits, characters per word, uppercase letters, lowercase letters, and more!
