---
title: "Example stuff"
author: "Andrew Heiss"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      # beforeInit: "libs/offline-mathjax.js"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")

# sass::sass(
#   sass::sass_file("css/ath-slides.scss"),
#   output = "css/ath-slides.css",
#   options = sass::sass_options(
#     # output_style = "compressed"
#   )
# )
```

```{r packages-data, include=FALSE}
library(tidyverse)
library(scales)
library(broom)
```

class: center middle main-title section-title-6

# Testing these<br>neat slides

.class-info[

**Session 1**

.light[PMAP 8521: Program Evaluation for Public Service<br>
Andrew Young School of Policy Studies<br>
Spring 2020]

]

---

class: title title-1

# Box test

.box-1[Something important]

.box-2[Something important]

.box-3[Something important]

.box-4[Something important]

.box-5[Something important]

.box-6[Something important]

.box-7[Something important]

---

class: title title-inv-1

# Inverted box test

.box-inv-1[Something important]

.box-inv-2[Something important]

.box-inv-3[Something important]

.box-inv-4[Something important]

.box-inv-5[Something important]

.box-inv-6[Something important]

.box-inv-7[Something important]

---

name: outline
class: title title-inv-6

# Plan for today

--

.box-4.medium[Endogeneity & exogeneity]

--

.box-5.medium[Instruments]

--

.box-2.medium[Using instruments]

--

.box-7.medium[IV with R]

--

.box-3.medium[Treatment effects & compliance]

???

*Important* notes go here.

---

class: center middle section-title section-title-4 animated fadeIn

# Endogeneity & exogeneity

---

layout: true
class: title title-4

---

# Does education cause higher earnings?

```{r echo=FALSE, fig.width=7, fig.height=2.5, out.width="85%"}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  labs(x = "Displacement")
```


.medium[
$$
\color{#FF851B}{\text{Earnings}_i} = \beta_0 + \beta_1 \color{#0074D9}{\text{Education}_i} + \epsilon_i
$$
]

---

# Split exogeneity and endogeneity

.box-inv-4[What if we could somehow separate education into its endogenous and exogenous parts?]

&nbsp; 

.SMALL[
$$
\begin{aligned}
\text{Earnings}_i =& \beta_0 + \beta_1 \text{Education}_i + \epsilon_i \\
& \beta_0 + \beta_1 (\text{Education}_i^\text{exog.} + \text{Education}_i^\text{endog.}) + \epsilon_i \\
& \beta_0 + \beta_1 \text{Education}_i^\text{exog.} + \underbrace{\beta_1 \text{Education}_i^{\text{endog.}} + \epsilon_i}_{w_i} \\
& \beta_0 + \beta_1 \text{Education}_i^\text{exog.} + w_i
\end{aligned}
$$
]

---

# Two columns with a title

.pull-left[

.box-5.medium.full-width[Stuff on this side]

]

.pull-right[

.box-5.medium[And more stuff on *this* side]

]

.box-inv-5[Where does this go?]

---

# Three columns with a title

.pull-left-3[

Something here.

]

.pull-middle-3[

Something here in the middle.

]

.pull-right-3[

Something else here.

]

---

# Slide with headings

## Subheading

.box-5[This is a slide with headings]

## Another subheading

.box-3[This too has a box]

### Level 3!!!

.box-6[Hey.]
